---
title: "Dashboard"
output: html_document
date: "2025-01-25"
---


## Welcome to the World of Pokemon
```{r}
library(datasets)
```

First lets bring in the Pokemon dataset.

```{r}

Pokemon = read.csv(file = 'data/Pokemon.csv')
head(Pokemon)
```

<br>
<br>

### Reasearch Question 1:

**What is the most common type for pokemon?**

Since there are secondary typing for Pokemon, we will have to combine the observations from the two variables.

<br>

But first lets see what the data shows for Pokemon that only have one type.

```{r}
# this is the Type.2 for charmander
# a Pokemon that has a singular "Fire" type

Pokemon$Type.2[5]
```
The data displays a "" or empty for a Pokemon that has to Type.2. This will muddy our data if we're looking for the most common type, since technically "" is a valid observation as far as the code is concerned.

<br>

These are the tables for Type.1 and Type.2 variables.

```{r}
table(Pokemon$Type.1)
table(Pokemon$Type.2)

# how many types are in Type.1
length(table(Pokemon$Type.1)) 

# how many types are in Type.2
length(table(Pokemon$Type.2))

```
As you can see there is an extra type in `table(Pokemon$Type.2)` that is "" with the number 386 to represent all of the observations/Pokemon with "" for Type.2.

<br>
This is the combined table of Type.1 and Type.2, minus the empty "" type Now we need to be careful, because this does not represent every Pokemon. As you can see when we look at the sum of the table, it shows 1214 when we only have 800 observations/Pokemon.
```{r}
combinedTypeTable = (table(Pokemon$Type.1) + table(Pokemon$Type.2)[-1])


paste("The sum is: ", sum(combinedTypeTable))
```


Finally lets see which type is the most common for pokemon, either as their Type.1 OR their Type.2.
```{r}
# finds the max(table), and then finds the index of the max value, then returns both the number and its name, in our case the type

combinedTypeTable[which(combinedTypeTable == max(combinedTypeTable))]
```

**As you can see, the most common typing for a Pokemon is Water.**

But as always a visualization will help in contextualizing this piece of information. 

```{r}
combinedTypeTable = sort(combinedTypeTable, decreasing=TRUE)
barplot(combinedTypeTable[1:5], main = "TOP 5 MOST COMMON TYPES IN POKEMON", ylab="Frequency", col=c("lightblue"))

```


<br>
<br>









### Reasearch Question 2: *ADD A "Total" Option to the Final Graph that Shows Total Amount of that Type

**What generation has the greatest number of each "type"(Type.1 and Type.2 combined) of Pokemon**

<br>

First we need to know how many different generations are present in our data.
```{r}
table(Pokemon$Generation)
```
There are 6 different generations of Pokemon listed in the data, with a varying number of Pokemon for each generation.


Now we need to parse the data so that we can see how many of each type is in each generation.

```{r}
# table for the number of each Type.1 type for all entries where the Generation is 1
table(Pokemon$Type.1[which(Pokemon$Generation == 1)])

# table for the number of each Type.2 type for all entries where the Generation is 1
table(Pokemon$Type.2[which(Pokemon$Generation == 1)])

length(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))
length(table(Pokemon$Type.2[which(Pokemon$Generation == 1)]))


```

But we run in to a problem due to the way that Pokemon types are set up. A Pokemon has one or two types, and the Type.2 is valued the same as Type.1. This means that we need to count both when looking at how many types of each Pokemon are in each generation. For example the first Pokemon in the data, Bulbasaur, would count as both a "Grass" and "Poison" type, with both holding equal weight. This causes discrepancies like we see above, where some types are missing from the Type.2 table and the Type.1 table . We will have to modify the table ourselves to make this work.

<br>

Since we will most likely have to do this for every generation, lets make a function that fills in the missing "types" and returns one vector with the combined values for both Type.1 and Type.2

```{r}
# this function will help us combine the two Type tables int o one table with all unique types from both Type.1 and Type.2, as well as add up duplicates


combineTypeTblByGen = function(generation){
  
  # merging the table of Type.1 and Type.2 for the generation
  mergedTypeTable = merge(table(Pokemon$Type.1[which(Pokemon$Generation == generation)]), table(Pokemon$Type.2[which(Pokemon$Generation == generation)])[-1], all= TRUE)
  
  
  # vector version of the merged tables above
  typeVec = mergedTypeTable[[2]]
  names(typeVec) = mergedTypeTable[[1]]
  
  
  
  # for loop that adds up the repeat values for the types, and then removes the duplicate name+value from the vector
  len = length(typeVec) - 1
  for(i in 1:len){
  
    if(i >= length(typeVec)){break}
  
    else if(names(typeVec[i]) == names(typeVec[i + 1])){

      typeVec[i+1] = typeVec[i] + typeVec[i+1]
      #print(typeVec[i])

      typeVec = typeVec[-c(i)]
  
    }

  }
  
  return(typeVec)
  
  
}


barplot(combineTypeTblByGen(1), las=2, main="Frequency of Types for Generation 1")


```

The plot above combines frequency of "types" in both Type1 and Type2 for generation 1, and because we have the function combineTypeTblByGen() we are able to replicate this for every other generation as well.


```{r, echo= F ,eval = FALSE}
#The code chunk below is the learning process / trying to figure out how to combine two tables of dinnerent dimensions together before I found the merge function





# figuring out how to add a vector with one value and name to a table
vec1 = 2
names(vec1) = c("Fire")
vec1

table(Pokemon$Type.1[which(Pokemon$Generation == 1)])

newtab = c(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1] , vec1)
newtab


# creating a vector with just the names of the types from the Type.1 and Type.2 tables
namesType1Vec = names(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))
namesType2Vec = names(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1])

is.vector(namesType2Vec)


# for loop that goes through each name in the namesType1Vec and checks if its == to any names in the namesType2Vec, and if so removes them from both vectors.

# this leaves us with only the types that are missing from both tables
# this solution has a time complexity of n^2, fix it later

for(type1 in namesType1Vec){
  
  for(type2 in namesType2Vec){
    if(type1 == type2){
      namesType1Vec = namesType1Vec[! namesType1Vec %in% type1]
      namesType2Vec = namesType2Vec[! namesType2Vec %in% type1]
      
    }
  }
  
}

namesType1Vec
namesType2Vec



# finding how to remove values from a vector -> hopefully can translate to table
vec2 = c("a", "b", "c")
vec2[! vec2 %in% c("a")]




# for Type.1 table
# creates a vector with the types missing from Type.1 with the types as names and values of 0 for all of them
MissingFromType1 = rep(0, times= length(namesType2Vec))
names(MissingFromType1) = c(namesType2Vec)

MissingFromType1


# for Type.2 table
# creates a vector with the types missing from Type.2 with the types as names and values of 0 for all of them
MissingFromType2 = rep(0, times= length(namesType1Vec))
names(MissingFromType2) = c(namesType1Vec)

MissingFromType2




Type1WithMissing = c(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]), MissingFromType1)

Type1WithMissing



Type2WithMissing = c(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1], MissingFromType2)

Type2WithMissing


class(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))


class(as.table(Type1WithMissing))




sort(as.table(Type1WithMissing)) 
sort(as.table(Type2WithMissing))

```



```{r, echo = F, eval = F}
# merging the table of Type.1 and Type.2 for Generation 1

mergedTypeTableGen1 = merge(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]), table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1], all= TRUE)

mergedTypeTableGen1


# vector version of the merged tables above

typeVec = mergedTypeTableGen1[[2]]
names(typeVec) = mergedTypeTableGen1[[1]]

typeVec
# for loop that adds up the repeat values for the types, and then removes the duplicate name+value from the vector
# 
# len = length(typeVec) - 1
# for(i in 1:len){
#   
#   if(i >= length(typeVec)){break}
#   
#   else if(names(typeVec[i]) == names(typeVec[i + 1])){
# 
#     typeVec[i+1] = typeVec[i] + typeVec[i+1]
#     print(typeVec[i])
# 
#     typeVec = typeVec[-c(i)]
# 
#   }
# 
# }
# 
# 
# 
# typeVec
# 
# length(typeVec)

```


<br>

Since we have the combined tables for every generation, we can move on to comparing the generations by Type. 

This next function will allow us to see how many of one specific type is in every generation. For example it will allow us to see the number of "Bug" types there are in generation 1 through 6.

```{r}

# function takes in a character argument that represents a Pokemon Type and returns a vector of length 6, with each index representing the number of that type of pokemon are in that generation number

compareTypeAcrossGen = function(type){
  
  oneTypeAllGensVec = c()
  
  for(i in 1:6){
    
    #print(combineTypeTblByGen(i)[type])
    
    oneTypeAllGensVec[i] = combineTypeTblByGen(i)[type]
   
    
  }

  names(oneTypeAllGensVec) = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")
  
  return(oneTypeAllGensVec)

}


compareTypeAcrossGen("Bug")

compareTypeAcrossGen("Dark")
```

```{r}

barplot(compareTypeAcrossGen("Bug"), main="Frequency of Bug Types in Each Generation(Gen)", las=2, col = "lightgreen")
```

We have the comparisons across generations for individual types, but that doesn't accomplish anything unless we want to create 18 different barplots. Instead lets use the plotly package to create an interactive barplot that shows the frequency of each type across the generations to get a broader understanding of the dataset.
```{r, echo = F, eval = F}
# Goal: Make a bar graph where I can toggle what Type is shown, and show the frequency of that Type for each generation.
# 
# What I need:
#   
#   - feed a dataframe into the plot_ly function, as well as other necessary arguments(which are labels, the type of plot, ect)
# 
#   - feed the graph into the add_trace function which allows you to add the different observations, for our case that would mean that originally the graph would only have "Bug" type, and then the add_trace() allows us to add "Dark" and "Dragon" types
# 
#   - feed the graph into the layout() function, which allows us to configure the layout of the graph. This is where the magic happens. Most important here is the (barmode) and (updatemenus) arguments. In the updatemenus, we need to choose thetype of menu we want. For this time it will be buttons
#
# Finally we can also turn off the displayModeBar at the top using the config() function

```

```{r, include = FALSE}
library(ggplot2)
library(plotly)
numPokemonTypes = 18



# setup for making traces, makes a list where the first element is a vector of the number of bug types in each generation and so on for every type 

types = names(combineTypeTblByGen(1))

typeByGenTraces = list()
for(i in 1:18){
  typeByGenTraces[[i]] = c(compareTypeAcrossGen(types[i]))
}
typeByGenTraces

# vector of colors to customize colors of the bars

colorsByPokemonType = c("lightgreen", "darkblue", "yellow", "pink", "brown", "orangered", "slateblue", "darkgreen", "khaki", "cyan", "lightgray", "purple", "hotpink", "tan", "deepskyblue","sienna", "skyblue", "lightsteelblue" )




# creates the dataframe for the plot_ly() function

Generations = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")
Bug = compareTypeAcrossGen("Bug")
data = data.frame(Generations, Bug)


# setting up the initial plot

q2FinalGraph = plot_ly(data, x = ~Generations, y = ~Bug, type = "bar", name = "Bug", marker = list(color = "lightgreen"))



# adding traces

for(i in 2:18){
  
  q2FinalGraph = q2FinalGraph %>% add_trace(y = typeByGenTraces[[i]], name = types[i], marker = list(color = colorsByPokemonType[i]))
  
  print(colorsByPokemonType[i])
}

  
# buttons and other configs  
  
q2FinalGraphButtons = list(
  # EX:
  # list(
  #   methods = "update",
  #   args = list(list(visible = c(TRUE, FALSE))),
  #   label = "BUG"
  #   
  # )
)

for(i in 1:numPokemonTypes){
  
  #makes a vector of length numPokemonTypes where vector[i] is TRUE, and every other index is false
  visibilityVector = c()
  
  if(i != 1){ visibilityVector[1:i-1] = FALSE }
  visibilityVector[i] = TRUE
  if(i != numPokemonTypes){ visibilityVector[i+1:18] = FALSE }
  
  
  #creates each individual button, with unique args and label
  q2FinalGraphButtons[[i]] = list(
    
    methods = "update",
    args = list(list(visible = visibilityVector)),
    label = names(combineTypeTblByGen(1)[i])
  )

}

# adds a button that toggles the graph to show every type again

q2FinalGraphButtons[[numPokemonTypes + 1]] = list(
  methods = "update",
  args = list(list(visible = rep(c(TRUE), times = 18))),
  label = "All"
)


# changes the layout of the graph and adds buttons to toggle between TYPES
q2FinalGraph = q2FinalGraph %>% layout(
  title = "Pokemon Type By Generation", 
  yaxis = list(title = "Count")
  ,barmode = "group",
  
  updatemenus = list(
    
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.5, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[1:12]
    ),
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.7, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[13:(numPokemonTypes + 1)]
    )
  )
)


# hides the tool bar that shows up on plot_ly interactive graphs
q2FinalGraph = q2FinalGraph %>% config(displayModeBar = FALSE)

q2FinalGraph

```

```{r}
typeByGenTraces[]
```



```{r}
q2FinalGraph

```

```{r, echo = F, eval = F}

# before i figured out how to iteratively add traces to plotly graphs
library(ggplot2)
library(plotly)
numPokemonTypes = 18

# creates the dataframe for the plot_ly() function
Generations = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")

# vector for the the first trace
Bug = compareTypeAcrossGen("Bug")

data = data.frame(Generations, Bug)


# making the graph and adding every trace, i should be doing this dynamically but it wasnt working soooo
q2FinalGraph = plot_ly(data, x = ~Generations, y = ~Bug, type = "bar", name = "BUG")




# creates the buttons for the graph. each list inside the main list represents one button and its attributes

q2FinalGraphButtons = list(
  # EX:
  # list(
  #   methods = "update",
  #   args = list(list(visible = c(TRUE, FALSE))),
  #   label = "BUG"
  #   
  # )
)

for(i in 1:numPokemonTypes){
  
  #makes a vector of length numPokemonTypes where vector[i] is TRUE, and every other index is false
  visibilityVector = c()
  
  if(i != 1){ visibilityVector[1:i-1] = FALSE }
  visibilityVector[i] = TRUE
  if(i != numPokemonTypes){ visibilityVector[i+1:18] = FALSE }
  
  
  #creates each individual button, with unique args and label
  q2FinalGraphButtons[[i]] = list(
    
    methods = "update",
    args = list(list(visible = visibilityVector)),
    label = names(combineTypeTblByGen(1)[i])
  )

}


# changes the layout of the graph and adds buttons to toggle between TYPES
q2FinalGraph = q2FinalGraph %>% layout(
  title = "Num of Each Pokemon Type in Every Generation", 
  yaxis = list(title = "Count")
  ,barmode = "group",
  
  updatemenus = list(
    
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.5, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[1:12]
    ),
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.7, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[13:18]
    )
  )
)


# hides the tool bar that shows up on plot_ly interactive graphs
q2FinalGraph = q2FinalGraph %>% config(displayModeBar = FALSE)

q2FinalGraph
```




## Lets do some Querying
```{r}

library(DBI)
library(RSQLite)
drv = dbDriver("SQLite")

pokemonDatabase = dbConnect(drv, dbname = "./data/veekun-pokedex.sqlite")
dbExecute(pokemonDatabase, "PRAGMA foreign_keys = on")

```


```{r}
dbListTables(pokemonDatabase)

```
This database and tables needs to be "cleaned" or organized into the information that I actually need --> practice organizing and putting data togehter with joins
\
Questions to ask/analyze:
\
how has the stats fro certain types changed over the course of the generartions?
\
is this change significant?
\

#### Goal 1: Create a table/df with all relevant info like pokemon speciest, pokedex number, stats, moves, pokedex flavor text, habitat, strengths and weaknesses, enocunters, locations, abilities

#### Goal 2:
Pokedex app? image, stats, and reads pokedex entry?
check normality for stats by type


hypothesis: flying types are significantly faster than ground, check?



### Research Question 3: How does each individual stat vary across the different types. NOTE: language_id of 9 is English
When I think of flying type Pokemon, I imagine fast aerial predators, so naturally you would think that their speed stat is significantly higher than that of ground/rock types that are usually large boulders and giants. But do the stats, get it?, actually back this up? **Lets Find Out!**
\


**Step 1: get a data frame with every flying type and their speed stat and every ground type and their speed stat**
\

Lets build this dataframe step by step.
\
**Step 1a: get a dataframe with every Pokemon and its speed stat**
```{r}
dbGetQuery(pokemonDatabase, 
           "PRAGMA FOREIGN_KEY_LIST(pokemon_stats)")

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, identifier, base_stat
           FROM pokemon_stats INNER JOIN stats
           ON pokemon_stats.stat_id = stats.id
           WHERE identifier = 'speed'
           LIMIT 20")

```


**Step 1b: dataframe with every pokemon's type**
```{r}
# dbGetQuery(pokemonDatabase, "
#            SELECT *
#            FROM pokemon_types")
# 
# dbGetQuery(pokemonDatabase, "
#            PRAGMA FOREIGN_KEY_LIST(pokemon_types)")
# 
# dbGetQuery(pokemonDatabase, "
#            SELECT *
#            FROM types")

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, slot, identifier as TYPE
           FROM pokemon_types INNER JOIN types
           ON pokemon_types.type_id = types.id
          LIMIT 20")

dbGetQuery(pokemonDatabase, "
           PRAGMA FOREIGN_KEY_LIST(pokemon_types)
           ")
```

**Step 1c: Now that we have two dataframes with every pokemon and its speed stat and another with every Pokemon and its type, we need a dataframe that combines these two.**
```{r}


```


NOTE: Join pokemon_types and types. pokemon_types gives pokemon id and their type id for slot 1 and 2. types gives the type names that correlate with the type id. 

```{r}
dbGetQuery(pokemonDatabase, "
           SELECT *
           FROM pokemon_types")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon_types)")




dbGetQuery(pokemonDatabase, "
           SELECT *
           FROM types")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(types)")


```







```{r}

dbDisconnect(pokemonDatabase)

```













































