---
title: "Dashboard"
output: html_document
date: "2025-01-25"
---


## Welcome to the World of Pokemon
```{r}
library(datasets)
```

First lets bring in the Pokemon dataset.

```{r}

Pokemon = read.csv(file = 'data/Pokemon.csv')
head(Pokemon)
```

<br>
<br>

### Reasearch Question 1:

**What is the most common type for pokemon?**

Since there are secondary typing for Pokemon, we will have to combine the observations from the two variables.

<br>

But first lets see what the data shows for Pokemon that only have one type.

```{r}
# this is the Type.2 for charmander
# a Pokemon that has a singular "Fire" type

Pokemon$Type.2[5]
```
The data displays a "" or empty for a Pokemon that has to Type.2. This will muddy our data if we're looking for the most common type, since technically "" is a valid observation as far as the code is concerned.

<br>

These are the tables for Type.1 and Type.2 variables.

```{r}
table(Pokemon$Type.1)
table(Pokemon$Type.2)

# how many types are in Type.1
length(table(Pokemon$Type.1)) 

# how many types are in Type.2
length(table(Pokemon$Type.2))

```
As you can see there is an extra type in `table(Pokemon$Type.2)` that is "" with the number 386 to represent all of the observations/Pokemon with "" for Type.2.

<br>
This is the combined table of Type.1 and Type.2, minus the empty "" type. Now we need to be careful, because this does not represent every Pokemon. As you can see when we look at the sum of the table, it shows 1214 when we only have 800 observations/Pokemon.
```{r}
combinedTypeTable = (table(Pokemon$Type.1) + table(Pokemon$Type.2)[-1])

combinedTypeTable
```

```{r}
paste("The sum is: ", sum(combinedTypeTable))
```


Finally lets see which type is the most common for pokemon, either as their Type.1 OR their Type.2.
```{r}
# finds the max(table), and then finds the index of the max value, then returns both the number and its name, in our case the type

combinedTypeTable[which(combinedTypeTable == max(combinedTypeTable))]
```

**As you can see, the most common typing for a Pokemon is Water.**

But as always a visualization will help in contextualizing this piece of information. 

```{r}
combinedTypeTable = sort(combinedTypeTable, decreasing=TRUE)
barplot(combinedTypeTable[1:5], main = "TOP 5 MOST COMMON TYPES IN POKEMON", ylab="Frequency", col=c("lightblue"))

```


<br>
<br>









### Reasearch Question 2: *ADD A "Total" Option to the Final Graph that Shows Total Amount of that Type

**What generation has the greatest number of each "type"(Type.1 and Type.2 combined) of Pokemon**

<br>

First we need to know how many different generations are present in our data.
```{r}
table(Pokemon$Generation)
```
There are 6 different generations of Pokemon listed in the data, with a varying number of Pokemon for each generation.


Now we need to parse the data so that we can see how many of each type is in each generation.

```{r}
# table for the number of each Type.1 type for all entries where the Generation is 1
table(Pokemon$Type.1[which(Pokemon$Generation == 1)])

# table for the number of each Type.2 type for all entries where the Generation is 1
table(Pokemon$Type.2[which(Pokemon$Generation == 1)])

length(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))
length(table(Pokemon$Type.2[which(Pokemon$Generation == 1)]))


```

But we run in to a problem due to the way that Pokemon types are set up. A Pokemon has one or two types, and the Type.2 is valued the same as Type.1. This means that we need to count both when looking at how many types of each Pokemon are in each generation. For example the first Pokemon in the data, Bulbasaur, would count as both a "Grass" and "Poison" type, with both holding equal weight. This causes discrepancies like we see above, where some types are missing from the Type.2 table and the Type.1 table . We will have to modify the table ourselves to make this work.

<br>

Since we will most likely have to do this for every generation, lets make a function that fills in the missing "types" and returns one vector with the combined values for both Type.1 and Type.2

```{r}
# this function will help us combine the two Type tables int o one table with all unique types from both Type.1 and Type.2, as well as add up duplicates


combineTypeTblByGen = function(generation){
  
  # merging the table of Type.1 and Type.2 for the generation
  mergedTypeTable = merge(table(Pokemon$Type.1[which(Pokemon$Generation == generation)]), table(Pokemon$Type.2[which(Pokemon$Generation == generation)])[-1], all= TRUE)
  
  
  # vector version of the merged tables above
  typeVec = mergedTypeTable[[2]]
  names(typeVec) = mergedTypeTable[[1]]
  
  
  
  # for loop that adds up the repeat values for the types, and then removes the duplicate name+value from the vector
  len = length(typeVec) - 1
  for(i in 1:len){
  
    if(i >= length(typeVec)){break}
  
    else if(names(typeVec[i]) == names(typeVec[i + 1])){

      typeVec[i+1] = typeVec[i] + typeVec[i+1]
      #print(typeVec[i])

      typeVec = typeVec[-c(i)]
  
    }

  }
  
  return(typeVec)
  
  
}


barplot(combineTypeTblByGen(1), las=2, main="Frequency of Types for Generation 1")


```

The plot above combines frequency of "types" in both Type1 and Type2 for generation 1, and because we have the function combineTypeTblByGen() we are able to replicate this for every other generation as well.


```{r, echo= F ,eval = FALSE}
#The code chunk below is the learning process / trying to figure out how to combine two tables of dinnerent dimensions together before I found the merge function





# figuring out how to add a vector with one value and name to a table
vec1 = 2
names(vec1) = c("Fire")
vec1

table(Pokemon$Type.1[which(Pokemon$Generation == 1)])

newtab = c(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1] , vec1)
newtab


# creating a vector with just the names of the types from the Type.1 and Type.2 tables
namesType1Vec = names(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))
namesType2Vec = names(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1])

is.vector(namesType2Vec)


# for loop that goes through each name in the namesType1Vec and checks if its == to any names in the namesType2Vec, and if so removes them from both vectors.

# this leaves us with only the types that are missing from both tables
# this solution has a time complexity of n^2, fix it later

for(type1 in namesType1Vec){
  
  for(type2 in namesType2Vec){
    if(type1 == type2){
      namesType1Vec = namesType1Vec[! namesType1Vec %in% type1]
      namesType2Vec = namesType2Vec[! namesType2Vec %in% type1]
      
    }
  }
  
}

namesType1Vec
namesType2Vec



# finding how to remove values from a vector -> hopefully can translate to table
vec2 = c("a", "b", "c")
vec2[! vec2 %in% c("a")]




# for Type.1 table
# creates a vector with the types missing from Type.1 with the types as names and values of 0 for all of them
MissingFromType1 = rep(0, times= length(namesType2Vec))
names(MissingFromType1) = c(namesType2Vec)

MissingFromType1


# for Type.2 table
# creates a vector with the types missing from Type.2 with the types as names and values of 0 for all of them
MissingFromType2 = rep(0, times= length(namesType1Vec))
names(MissingFromType2) = c(namesType1Vec)

MissingFromType2




Type1WithMissing = c(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]), MissingFromType1)

Type1WithMissing



Type2WithMissing = c(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1], MissingFromType2)

Type2WithMissing


class(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))


class(as.table(Type1WithMissing))




sort(as.table(Type1WithMissing)) 
sort(as.table(Type2WithMissing))

```



```{r, echo = F, eval = F}
# merging the table of Type.1 and Type.2 for Generation 1

mergedTypeTableGen1 = merge(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]), table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1], all= TRUE)

mergedTypeTableGen1


# vector version of the merged tables above

typeVec = mergedTypeTableGen1[[2]]
names(typeVec) = mergedTypeTableGen1[[1]]

typeVec
# for loop that adds up the repeat values for the types, and then removes the duplicate name+value from the vector
# 
# len = length(typeVec) - 1
# for(i in 1:len){
#   
#   if(i >= length(typeVec)){break}
#   
#   else if(names(typeVec[i]) == names(typeVec[i + 1])){
# 
#     typeVec[i+1] = typeVec[i] + typeVec[i+1]
#     print(typeVec[i])
# 
#     typeVec = typeVec[-c(i)]
# 
#   }
# 
# }
# 
# 
# 
# typeVec
# 
# length(typeVec)

```


<br>

Since we have the combined tables for every generation, we can move on to comparing the generations by Type. 

This next function will allow us to see how many of one specific type is in every generation. For example it will allow us to see the number of "Bug" types there are in generation 1 through 6.

```{r}

# function takes in a character argument that represents a Pokemon Type and returns a vector of length 6, with each index representing the number of that type of pokemon are in that generation number

compareTypeAcrossGen = function(type){
  
  oneTypeAllGensVec = c()
  
  for(i in 1:6){
    
    #print(combineTypeTblByGen(i)[type])
    
    oneTypeAllGensVec[i] = combineTypeTblByGen(i)[type]
   
    
  }

  names(oneTypeAllGensVec) = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")
  
  return(oneTypeAllGensVec)

}


compareTypeAcrossGen("Bug")

compareTypeAcrossGen("Dark")
```

```{r}

barplot(compareTypeAcrossGen("Bug"), main="Frequency of Bug Types in Each Generation(Gen)", las=2, col = "lightgreen")
```

We have the comparisons across generations for individual types, but that doesn't accomplish anything unless we want to create 18 different barplots. Instead lets use the plotly package to create an interactive barplot that shows the frequency of each type across the generations to get a broader understanding of the dataset.
```{r, echo = F, eval = F}
# Goal: Make a bar graph where I can toggle what Type is shown, and show the frequency of that Type for each generation.
# 
# What I need:
#   
#   - feed a dataframe into the plot_ly function, as well as other necessary arguments(which are labels, the type of plot, ect)
# 
#   - feed the graph into the add_trace function which allows you to add the different observations, for our case that would mean that originally the graph would only have "Bug" type, and then the add_trace() allows us to add "Dark" and "Dragon" types
# 
#   - feed the graph into the layout() function, which allows us to configure the layout of the graph. This is where the magic happens. Most important here is the (barmode) and (updatemenus) arguments. In the updatemenus, we need to choose thetype of menu we want. For this time it will be buttons
#
# Finally we can also turn off the displayModeBar at the top using the config() function

```

```{r, include = FALSE}
library(ggplot2)
library(plotly)
numPokemonTypes = 18



# setup for making traces, makes a list where the first element is a vector of the number of bug types in each generation and so on for every type 

types = names(combineTypeTblByGen(1))

typeByGenTraces = list()
for(i in 1:18){
  typeByGenTraces[[i]] = c(compareTypeAcrossGen(types[i]))
}
typeByGenTraces

# vector of colors to customize colors of the bars

colorsByPokemonType = c("lightgreen", "darkblue", "yellow", "pink", "brown", "orangered", "slateblue", "darkgreen", "khaki", "cyan", "lightgray", "purple", "hotpink", "tan", "deepskyblue","sienna", "skyblue", "lightsteelblue" )




# creates the dataframe for the plot_ly() function

Generations = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")
Bug = compareTypeAcrossGen("Bug")
data = data.frame(Generations, Bug)


# setting up the initial plot

q2FinalGraph = plot_ly(data, x = ~Generations, y = ~Bug, type = "bar", name = "Bug", marker = list(color = "lightgreen"))



# adding traces

for(i in 2:18){
  
  q2FinalGraph = q2FinalGraph %>% add_trace(y = typeByGenTraces[[i]], name = types[i], marker = list(color = colorsByPokemonType[i]))
  
  print(colorsByPokemonType[i])
}

  
# buttons and other configs  
  
q2FinalGraphButtons = list(
  # EX:
  # list(
  #   methods = "update",
  #   args = list(list(visible = c(TRUE, FALSE))),
  #   label = "BUG"
  #   
  # )
)

for(i in 1:numPokemonTypes){
  
  #makes a vector of length numPokemonTypes where vector[i] is TRUE, and every other index is false
  visibilityVector = c()
  
  if(i != 1){ visibilityVector[1:i-1] = FALSE }
  visibilityVector[i] = TRUE
  if(i != numPokemonTypes){ visibilityVector[i+1:18] = FALSE }
  
  
  #creates each individual button, with unique args and label
  q2FinalGraphButtons[[i]] = list(
    
    methods = "update",
    args = list(list(visible = visibilityVector)),
    label = names(combineTypeTblByGen(1)[i])
  )

}

# adds a button that toggles the graph to show every type again

q2FinalGraphButtons[[numPokemonTypes + 1]] = list(
  methods = "update",
  args = list(list(visible = rep(c(TRUE), times = 18))),
  label = "All"
)


# changes the layout of the graph and adds buttons to toggle between TYPES
q2FinalGraph = q2FinalGraph %>% layout(
  title = "Pokemon Type By Generation", 
  yaxis = list(title = "Count")
  ,barmode = "group",
  
  updatemenus = list(
    
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.5, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[1:12]
    ),
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.7, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[13:(numPokemonTypes + 1)]
    )
  )
)


# hides the tool bar that shows up on plot_ly interactive graphs
q2FinalGraph = q2FinalGraph %>% config(displayModeBar = FALSE)

q2FinalGraph

```

```{r}
typeByGenTraces[]
```



```{r}
q2FinalGraph

```

```{r, echo = F, eval = F}

# before i figured out how to iteratively add traces to plotly graphs
library(ggplot2)
library(plotly)
numPokemonTypes = 18

# creates the dataframe for the plot_ly() function
Generations = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")

# vector for the the first trace
Bug = compareTypeAcrossGen("Bug")

data = data.frame(Generations, Bug)


# making the graph and adding every trace, i should be doing this dynamically but it wasnt working soooo
q2FinalGraph = plot_ly(data, x = ~Generations, y = ~Bug, type = "bar", name = "BUG")




# creates the buttons for the graph. each list inside the main list represents one button and its attributes

q2FinalGraphButtons = list(
  # EX:
  # list(
  #   methods = "update",
  #   args = list(list(visible = c(TRUE, FALSE))),
  #   label = "BUG"
  #   
  # )
)

for(i in 1:numPokemonTypes){
  
  #makes a vector of length numPokemonTypes where vector[i] is TRUE, and every other index is false
  visibilityVector = c()
  
  if(i != 1){ visibilityVector[1:i-1] = FALSE }
  visibilityVector[i] = TRUE
  if(i != numPokemonTypes){ visibilityVector[i+1:18] = FALSE }
  
  
  #creates each individual button, with unique args and label
  q2FinalGraphButtons[[i]] = list(
    
    methods = "update",
    args = list(list(visible = visibilityVector)),
    label = names(combineTypeTblByGen(1)[i])
  )

}


# changes the layout of the graph and adds buttons to toggle between TYPES
q2FinalGraph = q2FinalGraph %>% layout(
  title = "Num of Each Pokemon Type in Every Generation", 
  yaxis = list(title = "Count")
  ,barmode = "group",
  
  updatemenus = list(
    
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.5, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[1:12]
    ),
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.7, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[13:18]
    )
  )
)


# hides the tool bar that shows up on plot_ly interactive graphs
q2FinalGraph = q2FinalGraph %>% config(displayModeBar = FALSE)

q2FinalGraph
```




## Lets do some Querying
```{r}

library(DBI)
library(RSQLite)
drv = dbDriver("SQLite")

pokemonDatabase = dbConnect(drv, dbname = "./data/veekun-pokedex.sqlite")
dbExecute(pokemonDatabase, "PRAGMA foreign_keys = on")

```


```{r, eval = FALSE}
dbListTables(pokemonDatabase)

```

### Research Question 3: How does each individual stat vary across the different types. 
When I think of flying type Pokemon, I imagine fast aerial predators, so naturally you would think that their speed stat is significantly higher than that of ground/rock types that are usually large boulders and giants. But do the stats, get it?, actually back this up? **Lets Find Out!**
\


**Step 1: get a data frame with every flying type and their speed stat and every ground type and their speed stat**
\

Lets build this dataframe step by step.
\
**Step 1a: get a dataframe with every Pokemon and its speed stat**
```{r}
dbGetQuery(pokemonDatabase, 
           "PRAGMA FOREIGN_KEY_LIST(pokemon_stats)")

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, identifier, base_stat
           FROM pokemon_stats INNER JOIN stats
           ON pokemon_stats.stat_id = stats.id
           WHERE identifier = 'speed'
           LIMIT 20")

```


**Step 1b: dataframe with every pokemon's type**
```{r}
# dbGetQuery(pokemonDatabase, "
#            SELECT *
#            FROM pokemon_types")
# 
# dbGetQuery(pokemonDatabase, "
#            PRAGMA FOREIGN_KEY_LIST(pokemon_types)")
# 
# dbGetQuery(pokemonDatabase, "
#            SELECT *
#            FROM types")
dbGetQuery(pokemonDatabase, "
           PRAGMA FOREIGN_KEY_LIST(pokemon_types)
           ")  

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, slot, identifier as TYPE
           FROM pokemon_types INNER JOIN types
           ON pokemon_types.type_id = types.id
          LIMIT 20")


```

**Step 1c: Now that we have two dataframes with every pokemon and its speed stat and another with every Pokemon and its type, we need a dataframe that combines these two.**
```{r, eval = FALSE, echo = FALSE}

dbGetQuery(pokemonDatabase, "SELECT * FROM pokemon_stats")

dbGetQuery(pokemonDatabase, "SELECT * FROM stats")

dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon_stats)")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon)")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(stats)")


```

```{r}
# initial query with all required columns
dbGetQuery(pokemonDatabase, "
           SELECT pokemon.id, pokemon.identifier, pokemon_types.pokemon_id, pokemon_types.type_id, types.identifier, pokemon_stats.stat_id, pokemon_stats.base_stat, stats.identifier
           
           FROM pokemon_types

           INNER JOIN pokemon ON pokemon_types.pokemon_id = pokemon.id
           INNER JOIN types ON pokemon_types.type_id = types.id
           INNER JOIN pokemon_stats ON pokemon_stats.pokemon_id = pokemon.id
           INNER JOIN stats ON pokemon_stats.stat_id = stats.id
           LIMIT 20
           ")

```


**Step 1d: Save the dataframe so that we can do statistics on it with R**
```{r}

# cleaned up query with all required info for flying and ground type pokemon
flynGrndSPD_df = dbGetQuery(pokemonDatabase, "
           SELECT pokemon.identifier AS name, types.identifier AS type, stats.identifier AS stat_name, pokemon_stats.base_stat
           
           FROM pokemon_types

           INNER JOIN pokemon ON pokemon_types.pokemon_id = pokemon.id
           INNER JOIN types ON pokemon_types.type_id = types.id
           INNER JOIN pokemon_stats ON pokemon_stats.pokemon_id = pokemon.id
           INNER JOIN stats ON pokemon_stats.stat_id = stats.id
            
           WHERE stat_name LIKE 'speed' AND (type LIKE 'flying' OR type LIKE 'ground')
           
           ")

# data frame with flying types and their speed stats
flyingSpeeds = flynGrndSPD_df[which(flynGrndSPD_df$type == "flying"), ]

# data frame with ground types and their speed stats
groundSpeeds = flynGrndSPD_df[which(flynGrndSPD_df$type == "ground"), ]

# first 10 entries
flynGrndSPD_df[1:10,]
```


**Step 2: Determine normality**
\
In order to figure out what hypothesis test to choose, lets figure out if our data follows a normal distribution.
```{r}
hist(flyingSpeeds$base_stat, main = "Flying Type Speed Distribution", xlab = "Speed stat", breaks=20)
```

```{r}
hist(groundSpeeds$base_stat, main = "Ground Type Speed Distribution", xlab = "Speed stat", breaks = 20)
```

```{r}
qqnorm(flyingSpeeds$base_stat, main  = "Normal Q-Q Plot for Flying Types")
qqline(flyingSpeeds$base_stat, col = "red")

```

```{r}
qqnorm(groundSpeeds$base_stat, main  = "Normal Q-Q Plot for Ground Types")
qqline(groundSpeeds$base_stat, col = "red")
```


While the data doesnt appear to be normal, it helps to do a formal test to make sure that we are making an accurate interpretation of our data that isn't solely reliant on visuals.
\
In order to do that, we'll conduct a **Shapiro Wilk Test** on a random sample from both groups.
\
**Step 2b: conduct Shapiro Wilk Test**
\
First we need to take random samples from both groups with sample size n = 30.
```{r}
set.seed(923238)
flySpeedSample = sample(flyingSpeeds$base_stat, 30, replace = TRUE)
grndSpeedSample = sample(groundSpeeds$base_stat, 30, replace = TRUE)
```

Then we can use R's built in function for the Shapiro test for normality.

```{r}
shapiro.test(flySpeedSample)
shapiro.test(grndSpeedSample)
```
Lets interpret these results: For info, the Shapiro Wilk test is a hypothesis test where the null hypothesis is that a random sample comes from a normally distributed dataset.
\
\
The W statistic is the result of the Shapiro Wilk Test, which measures how  close the sample matches a normal distribution, with 1 being normal and 0 being not normal. The P value determines if we can reject the null hypothesis that the sample comes from a normally distributed dataset. If the p value result is <= 0.05, we can reject the null hypothesis, however if the p value result is > 0.05 we are unable to reject the null hypothesis that the random sample comes from a normally distributed dataset.
\
\
However, we encounter a problem. Depending on the random sample, the samples either fail or pass the test. We want to see if it is passing this test a reliable number of times. So lets build a function that does the shapiro wilk test on the 100 different samples, and lets have a p value of 0.05, ie if less that 95% of the samples fail the test, then we can reject the null hypothesis that the samples come from a normally distributed population.

```{r}
shapiroSample = function(dataset, reps, sampleSize){
  

  pass_vec = c()
  for(i in 1:reps){
    sample = sample(dataset, sampleSize, replace = TRUE)
    outcome = shapiro.test(sample)
    
    # checks to see if this sample rejects the null hypothesis or not
    # if it does adds a 0 for fail, and if not adds a 1 for pass
    
    if(outcome[2] > 0.05){pass_vec[i] = 1}
    else if(outcome[2] <= 0.05){pass_vec[i] = 0}
    
  }
  
  return(pass_vec)
}

```


Now lets execute this function for our data.
```{r}
set.seed(2354)
mean(shapiroSample(flyingSpeeds$base_stat, 100, 30))

mean(shapiroSample(groundSpeeds$base_stat, 100, 30))
```
Since the percentage of samples out of 100 that pass the Shapiro Wilk test are nowhere close to 95% for both flying and ground types, we can say that our data is not normally distributed.
\
\
**So what do we do now?**
\
We need to find an appropriate statistical test that will give us the findings we are looking for without normality of the data being a requirement.
\
We will use the **MANN-WHITNEY U TEST** which will allow us to see if our sample of flying and ground type Pokemons' speeds come from significantly different populations, ie if the populations have a significant difference in their distribution.
\
**Step 3: Conduct Mann-Whitney U Test**
\
**Step 3a: Rank the speed stats of both samples**
```{r, echo = FALSE}
# samples of flying that we used for the shapiro wilk test
flySpeedSample
grndSpeedSample

# creates a data frame with columns type and speed_stat and populates them with the samples
fly_grnd_speed_ranked = data.frame(rep(c("flying", "ground"), each = length(flySpeedSample)), c(flySpeedSample, grndSpeedSample) )
names(fly_grnd_speed_ranked) = c("type", "speed_stat")

# orders the speed_stat column(ascending)
fly_grnd_speed_ranked = fly_grnd_speed_ranked[order(fly_grnd_speed_ranked$speed_stat), ]

# adds a column with the ranks 
fly_grnd_speed_ranked$rank = 1:dim(fly_grnd_speed_ranked)[1]
fly_grnd_speed_ranked
```

**Step 3b: Separate the flying and ground types to get their rank sums**
```{r}
ground_speed_ranks = fly_grnd_speed_ranked[which(fly_grnd_speed_ranked$type == "ground"), ]

flying_speed_ranks = fly_grnd_speed_ranked[which(fly_grnd_speed_ranked$type == "flying"), ]
```
**Step 3c: calculate the rank sums for both types **

```{r}
ground_speed_ranksum = sum(ground_speed_ranks$rank)
flying_speed_ranksum = sum(flying_speed_ranks$rank)
ground_speed_ranksum
flying_speed_ranksum
```


**Step 3d: Determine the U statistic and the P value**

The U statistic tells you how often values from one of the samples rank below the other. U1 is the number of times values from the first sample rank above values from the second one. U2 is vice versa. The U statistic is min(U1, U2). However this is for small sample sizes. 
\
When sample sizes n > 20, U statistic still shows how much overlap there is between the two sample ranks. However as the sample size becomes sufficiently large, the distribution of U, ie all of the possible values that U could be, becomes approximately normal.
\
This means two things: that the U table is no longer helpul as there are too many possible values that U can be, and that we have to calculate the z-score.
\
The z-score tells you how many standard deviations your calculated U statistic is from the mean U statistic.
\
Then we can use pnorm(z-score) to get the p value which tells you how likely, given the null hypothesis is true, that you get that specific z-score. 
```{r}
n1 = dim(ground_speed_ranks)[1]
n2 = dim(flying_speed_ranks)[1]

U1 = ( (n1 * n2) + ((n1 * (n1+1)) / 2) ) - ground_speed_ranksum
U2 = ( (n1 * n2) + ((n1 * (n1+1)) / 2) ) - flying_speed_ranksum

U1
U2

Ustatistic = min(U1, U2)
muU = (n1 * n2) / 2
sigmaU = sqrt(((n1 * n2) * (n1 + n2 + 1)) / 12)
  
zscore = (Ustatistic - muU) / sigmaU

zscore

# gives the p value
pnorm(zscore)
```
Since our p value is extremely small, much much smaller than our chosen significance value of 0.05, we can confidently reject the null hypothesis that there is no significant difference between the speed stats of flying and ground type Pokemon. 
\
And since our z-score was negative we can colclude the direction, ie that the first group(ground types) had lower ranks.
\
##### Colclusion:
Putting all of this information together, we can conclude that Flying type and Ground type Pokemon have significantly different speed stats and furthermore, Flying types do indeed have significantly higher speed stats than Groudn types.





#### Research Question 4: are pokemon with certain abilities statistically weaker than pokemon with certain other abilities, ie are abilites statistically related to base stats in some way, and if so what does this tell us?

### NOTE: USE VIEWS IF THERE IS A QUERY YOU NEED TO KEEP MAKING
What insight do we gain from the existence/nonexistence of a relationship between base stats and abilities?
\

Step 1: Lets query to get a dataframe with Pokemon, their name, their abilities, and their base stat totals.


```{r, eval = FALSE}
# querying the relations that have the fields for the final data frame

dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon_abilities)")

dbGetQuery(pokemonDatabase, "
           SELECT *
           FROM pokemon_abilities
           ")

dbGetQuery(pokemonDatabase, "
           SELECT *
           FROM abilities")


dbGetQuery(pokemonDatabase, "
           SELECT *
           FROM pokemon_stats")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon_stats)")


# getting the total base stat of each pokemon
dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, SUM(base_stat)
           FROM pokemon_stats
           GROUP BY pokemon_id
           ")


```



```{r}

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_abilities.pokemon_id, pokemon.identifier, pokemon_abilities.ability_id, abilities.identifier, SUM(pokemon_stats.base_stat)
           FROM pokemon_abilities
           
           INNER JOIN abilities ON pokemon_abilities.ability_id = abilities.id
           INNER JOIN pokemon ON pokemon_abilities.pokemon_id = pokemon.id
           INNER JOIN pokemon_stats ON pokemon.id = pokemon_stats.pokemon_id
           
           GROUP BY pokemon_abilities.pokemon_id, abilities.identifier
           LIMIT 25")



```

Now that we have all of the pokemon, their abilities, and base stats, lets see how many abilities and how many of each one there are.

```{r}
ability_nums = dbGetQuery(pokemonDatabase, "
           SELECT DISTINCT ability_id 
           FROM pokemon_abilities")
dim(ability_nums)[1]
# 
dbGetQuery(pokemonDatabase, "
           SELECT ability_id, COUNT(ability_id)
           FROM pokemon_abilities
           GROUP BY ability_id
           ")
```



#### Research Question X: 
Based on historical data, can we build a predictive model that estimates a Pokémon’s base stats based on its type, weight, height, and generation
```{r}

```




```{r}

dbDisconnect(pokemonDatabase)

```













































