---
title: "Pokémon Data Analysis"
output: html_document
---
<!-- <style> -->
<!--   body { -->
<!--     margin-left: 300px; -->
<!--     margin-right: 300px; -->
<!--   } -->
<!-- </style> -->
<br>
<nav style="background-color:#f8f9fa; padding: 10px; margin-bottom: 20px; border-bottom: 1px solid #ddd;"> <a href="#common" style="margin-right: 20px;">Common Types</a> <a href="#frequency" style="margin-right: 20px;">Type Frequency</a> <a href="#hypothesis-testing" style="margin-right: 20px;">Hypothesis Testing</a></nav>
<br>

### **Overview**
Analysis of Pokémon Type Trends and inter-Type Speed Comparison
<br>
This project explores patterns in Pokémon types across the first six generations and examines potential statistical differences in base Speed stats between Flying and Ground types. The analysis was conducted using a csv containing information about each Pokémon's name, type(s), base stats, and generation. The hypothesis testing was conducted using a SQlite database. The project is divided into three key components:

<ol>
  <li><div><h5>**Most Common Pokémon Types Across Generations**</h5><p>Analysis to identify the most commonly occurring Pokémon types over all 6 generations. This reveals trends in game design and type distribution for the franchise.</p></div>
  </li>
  <li><div><h5>**Type Frequency by Generation**</h5><p>To further understand how type popularity evolves, the distribution types were analyzed  for each generation and each type. Visualizations such as static and interactive bar plots were used to illustrate shifts and patterns.</p></div>
  </li>
  <li><div><h5>**Hypothesis Testing: Speed Comparison Between Flying and Ground Types**</h5><p>A statistical hypothesis test (Mann-Whitney U test) was conducted to assess whether Flying type Pokémon have significantly higher base Speed stats than Ground type Pokémon. The null hypothesis assumes no difference in the distributions of Speed stats between the Pokémon  of the two groups, while the alternative hypothesis states that Flying types’ distribution rank significantly higher than Ground types when the two are put together. The results provide insights into how type might relate to base stat design decisions.</p></div></li>
</ol>  

 
This project demonstrates the use of R, Plotly, and SQLite for data wrangling, visualization, and statistical analysis, with a focus on uncovering meaningful insights from data.


```{r}
library(datasets)
```


```{r}

Pokemon = read.csv(file = 'data/Pokemon.csv')
head(Pokemon)
```

<br>

<a id="common"></a>

### Exploratory Data Analysis

##### **What is the most common type of Pokémon?**

Need to address missing values in Type.2 since some Pokémon only have 1 type.
<br>
What the data shows for Pokemon that only have one type
```{r}
# this is the Type.2 for charmander
# a Pokemon that has a singular "Fire" type

Pokemon$Type.2[5]
```
The data displays a "" or empty for a Pokémon that has no Type.2. This will muddy the data when looking for the most common type.
<br>

Tables for Type.1 and Type.2 variables.
```{r}
table(Pokemon$Type.1)
table(Pokemon$Type.2)

# how many types are in Type.1
length(table(Pokemon$Type.1)) 

# how many types are in Type.2
length(table(Pokemon$Type.2))

```

There is an extra type in `table(Pokemon$Type.2)` that is "" with the number 386 to represent all of the observations/Pokémon with "" for Type.2.

<br>
This is the combined table of Type.1 and Type.2, minus the empty "" type. However this does not represent every Pokemon. The sum of the table shows 1214 when we only have 800 observations/Pokémon.

<br>
This 1214 occurs because it is counting a Pokémon like Bulbasaur(grass/poison) twice, once as a grass type and again as a poison type, but for the purposes of finding the most common type thats okay.
```{r}
combinedTypeTable = (table(Pokemon$Type.1) + table(Pokemon$Type.2)[-1])

combinedTypeTable
```

```{r}
paste("The sum is: ", sum(combinedTypeTable))
```


The max value in the table and the name associated with it.
```{r}
# finds the max(table), and then finds the index of the max value, then returns both the number and its name, in our case the type

combinedTypeTable[which(combinedTypeTable == max(combinedTypeTable))]
```

**The most common type for a Pokémon is Water.**
```{r}
combinedTypeTable = sort(combinedTypeTable, decreasing=TRUE)
barplot(combinedTypeTable[1:5], main = "TOP 5 MOST COMMON TYPES IN POKEMON", ylab="Frequency", col=c("lightblue"))

```


<br>
<br>

##### **How many of each type of Pokémon is there in every Generation?**

How many different generations are present in the data
```{r}
table(Pokemon$Generation)
```
There are 6 different generations of Pokémon listed in the data, with a varying number of Pokemon for each generation.

<br>
Parse the data to see how many of each type is in Generation 1.

```{r}
# table for the number of each Type.1 type for all entries where the Generation is 1
table(Pokemon$Type.1[which(Pokemon$Generation == 1)])

# table for the number of each Type.2 type for all entries where the Generation is 1
table(Pokemon$Type.2[which(Pokemon$Generation == 1)])

length(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))
length(table(Pokemon$Type.2[which(Pokemon$Generation == 1)]))


```

But due to the way that Pokémon types are set up there is an issue. A Pokémon has one or two types, and the Type.2 is valued the same as Type.1. This causes discrepancies like above, where some types are missing from the Type.2 table and the Type.1 table . 

<br>
Make a function that fills in the missing "types" and returns one vector with the combined values for both Type.1 and Type.2

```{r}
# this function will help us combine the two Type tables int o one table with all unique types from both Type.1 and Type.2, as well as add up duplicates


combineTypeTblByGen = function(generation){
  
  # merging the table of Type.1 and Type.2 for the generation
  mergedTypeTable = merge(table(Pokemon$Type.1[which(Pokemon$Generation == generation)]), table(Pokemon$Type.2[which(Pokemon$Generation == generation)])[-1], all= TRUE)
  
  
  # vector version of the merged tables above
  typeVec = mergedTypeTable[[2]]
  names(typeVec) = mergedTypeTable[[1]]
  
  
  
  # for loop that adds up the repeat values for the types, and then removes the duplicate name+value from the vector
  len = length(typeVec) - 1
  for(i in 1:len){
  
    if(i >= length(typeVec)){break}
  
    else if(names(typeVec[i]) == names(typeVec[i + 1])){

      typeVec[i+1] = typeVec[i] + typeVec[i+1]
      #print(typeVec[i])

      typeVec = typeVec[-c(i)]
  
    }

  }
  
  return(typeVec)
  
  
}


barplot(combineTypeTblByGen(1), las=2, main="Frequency of Types for Generation 1", col=c("lightblue"))


```

The plot above combines frequency of "types" in both Type1 and Type2 for generation 1, and the function combineTypeTblByGen() replicate this for every other generation.


```{r, echo= F ,eval = FALSE}
#The code chunk below is the learning process / trying to figure out how to combine two tables of dinnerent dimensions together before I found the merge function





# figuring out how to add a vector with one value and name to a table
vec1 = 2
names(vec1) = c("Fire")
vec1

table(Pokemon$Type.1[which(Pokemon$Generation == 1)])

newtab = c(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1] , vec1)
newtab


# creating a vector with just the names of the types from the Type.1 and Type.2 tables
namesType1Vec = names(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))
namesType2Vec = names(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1])

is.vector(namesType2Vec)


# for loop that goes through each name in the namesType1Vec and checks if its == to any names in the namesType2Vec, and if so removes them from both vectors.

# this leaves us with only the types that are missing from both tables
# this solution has a time complexity of n^2, fix it later

for(type1 in namesType1Vec){
  
  for(type2 in namesType2Vec){
    if(type1 == type2){
      namesType1Vec = namesType1Vec[! namesType1Vec %in% type1]
      namesType2Vec = namesType2Vec[! namesType2Vec %in% type1]
      
    }
  }
  
}

namesType1Vec
namesType2Vec



# finding how to remove values from a vector -> hopefully can translate to table
vec2 = c("a", "b", "c")
vec2[! vec2 %in% c("a")]




# for Type.1 table
# creates a vector with the types missing from Type.1 with the types as names and values of 0 for all of them
MissingFromType1 = rep(0, times= length(namesType2Vec))
names(MissingFromType1) = c(namesType2Vec)

MissingFromType1


# for Type.2 table
# creates a vector with the types missing from Type.2 with the types as names and values of 0 for all of them
MissingFromType2 = rep(0, times= length(namesType1Vec))
names(MissingFromType2) = c(namesType1Vec)

MissingFromType2




Type1WithMissing = c(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]), MissingFromType1)

Type1WithMissing



Type2WithMissing = c(table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1], MissingFromType2)

Type2WithMissing


class(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]))


class(as.table(Type1WithMissing))




sort(as.table(Type1WithMissing)) 
sort(as.table(Type2WithMissing))

```



```{r, echo = F, eval = F}
# merging the table of Type.1 and Type.2 for Generation 1

mergedTypeTableGen1 = merge(table(Pokemon$Type.1[which(Pokemon$Generation == 1)]), table(Pokemon$Type.2[which(Pokemon$Generation == 1)])[-1], all= TRUE)

mergedTypeTableGen1


# vector version of the merged tables above

typeVec = mergedTypeTableGen1[[2]]
names(typeVec) = mergedTypeTableGen1[[1]]

typeVec
# for loop that adds up the repeat values for the types, and then removes the duplicate name+value from the vector
# 
# len = length(typeVec) - 1
# for(i in 1:len){
#   
#   if(i >= length(typeVec)){break}
#   
#   else if(names(typeVec[i]) == names(typeVec[i + 1])){
# 
#     typeVec[i+1] = typeVec[i] + typeVec[i+1]
#     print(typeVec[i])
# 
#     typeVec = typeVec[-c(i)]
# 
#   }
# 
# }
# 
# 
# 
# typeVec
# 
# length(typeVec)

```

This next function outputs a vector that shows how many of one specific type is in every Generation. For example it'll show the number of "Bug" types there are in Generation 1 through 6.

```{r}

# function takes in a character argument that represents a Pokemon Type and returns a vector of length 6, with each index representing the number of that type of pokemon are in that generation number

compareTypeAcrossGen = function(type){
  
  oneTypeAllGensVec = c()
  
  for(i in 1:6){
    
    #print(combineTypeTblByGen(i)[type])
    
    oneTypeAllGensVec[i] = combineTypeTblByGen(i)[type]
   
    
  }

  names(oneTypeAllGensVec) = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")
  
  return(oneTypeAllGensVec)

}


compareTypeAcrossGen("Bug")

compareTypeAcrossGen("Dark")
```

```{r}

barplot(compareTypeAcrossGen("Bug"), main="Frequency of Bug Types in Each Generation(Gen)", las=2, col = "lightgreen")
```

**Use the Plotly package to create an interactive barplot that shows the frequency of each type across the 6 Generations**
```{r, echo = F, eval = F}
# Goal: Make a bar graph where I can toggle what Type is shown, and show the frequency of that Type for each generation.
# 
# What I need:
#   
#   - feed a dataframe into the plot_ly function, as well as other necessary arguments(which are labels, the type of plot, ect)
# 
#   - feed the graph into the add_trace function which allows you to add the different observations, for our case that would mean that originally the graph would only have "Bug" type, and then the add_trace() allows us to add "Dark" and "Dragon" types
# 
#   - feed the graph into the layout() function, which allows us to configure the layout of the graph. This is where the magic happens. Most important here is the (barmode) and (updatemenus) arguments. In the updatemenus, we need to choose thetype of menu we want. For this time it will be buttons
#
# Finally we can also turn off the displayModeBar at the top using the config() function

```

```{r, include = FALSE}
library(ggplot2)
library(plotly)
numPokemonTypes = 18



# setup for making traces, makes a list where the first element is a vector of the number of bug types in each generation and so on for every type 

types = names(combineTypeTblByGen(1))

typeByGenTraces = list()
for(i in 1:18){
  typeByGenTraces[[i]] = c(compareTypeAcrossGen(types[i]))
}
typeByGenTraces

# vector of colors to customize colors of the bars

colorsByPokemonType = c("lightgreen", "darkblue", "yellow", "pink", "brown", "orangered", "slateblue", "darkgreen", "khaki", "cyan", "lightgray", "purple", "hotpink", "tan", "deepskyblue","sienna", "skyblue", "lightsteelblue" )




# creates the dataframe for the plot_ly() function

Generations = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")
Bug = compareTypeAcrossGen("Bug")
data = data.frame(Generations, Bug)


# setting up the initial plot

q2FinalGraph = plot_ly(data, x = ~Generations, y = ~Bug, type = "bar", name = "Bug", marker = list(color = "lightgreen"))



# adding traces

for(i in 2:18){
  
  q2FinalGraph = q2FinalGraph %>% add_trace(y = typeByGenTraces[[i]], name = types[i], marker = list(color = colorsByPokemonType[i]))
  
  print(colorsByPokemonType[i])
}

  
# buttons and other configs  
  
q2FinalGraphButtons = list(
  # EX:
  # list(
  #   methods = "update",
  #   args = list(list(visible = c(TRUE, FALSE))),
  #   label = "BUG"
  #   
  # )
)

for(i in 1:numPokemonTypes){
  
  #makes a vector of length numPokemonTypes where vector[i] is TRUE, and every other index is false
  visibilityVector = c()
  
  if(i != 1){ visibilityVector[1:i-1] = FALSE }
  visibilityVector[i] = TRUE
  if(i != numPokemonTypes){ visibilityVector[i+1:18] = FALSE }
  
  
  #creates each individual button, with unique args and label
  q2FinalGraphButtons[[i]] = list(
    
    methods = "update",
    args = list(list(visible = visibilityVector)),
    label = names(combineTypeTblByGen(1)[i])
  )

}

# adds a button that toggles the graph to show every type again

q2FinalGraphButtons[[numPokemonTypes + 1]] = list(
  methods = "update",
  args = list(list(visible = rep(c(TRUE), times = 18))),
  label = "All"
)


# changes the layout of the graph and adds buttons to toggle between TYPES
q2FinalGraph = q2FinalGraph %>% layout(
  title = "Pokemon Type By Generation (interactive)", 
  yaxis = list(title = "Count")
  ,barmode = "group",
  
  updatemenus = list(
    
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.5, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[1:12]
    ),
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.7, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[13:(numPokemonTypes + 1)]
    )
  )
)


# hides the tool bar that shows up on plot_ly interactive graphs
q2FinalGraph = q2FinalGraph %>% config(displayModeBar = FALSE)

q2FinalGraph

```

```{r}
typeByGenTraces[]
```

<a id="frequency"></a>

```{r}
q2FinalGraph

```

```{r, echo = F, eval = F}

# before i figured out how to iteratively add traces to plotly graphs
library(ggplot2)
library(plotly)
numPokemonTypes = 18

# creates the dataframe for the plot_ly() function
Generations = c("Gen 1", "Gen 2", "Gen 3", "Gen 4", "Gen 5", "Gen 6")

# vector for the the first trace
Bug = compareTypeAcrossGen("Bug")

data = data.frame(Generations, Bug)


# making the graph and adding every trace, i should be doing this dynamically but it wasnt working soooo
q2FinalGraph = plot_ly(data, x = ~Generations, y = ~Bug, type = "bar", name = "BUG")




# creates the buttons for the graph. each list inside the main list represents one button and its attributes

q2FinalGraphButtons = list(
  # EX:
  # list(
  #   methods = "update",
  #   args = list(list(visible = c(TRUE, FALSE))),
  #   label = "BUG"
  #   
  # )
)

for(i in 1:numPokemonTypes){
  
  #makes a vector of length numPokemonTypes where vector[i] is TRUE, and every other index is false
  visibilityVector = c()
  
  if(i != 1){ visibilityVector[1:i-1] = FALSE }
  visibilityVector[i] = TRUE
  if(i != numPokemonTypes){ visibilityVector[i+1:18] = FALSE }
  
  
  #creates each individual button, with unique args and label
  q2FinalGraphButtons[[i]] = list(
    
    methods = "update",
    args = list(list(visible = visibilityVector)),
    label = names(combineTypeTblByGen(1)[i])
  )

}


# changes the layout of the graph and adds buttons to toggle between TYPES
q2FinalGraph = q2FinalGraph %>% layout(
  title = "Num of Each Pokemon Type in Every Generation", 
  yaxis = list(title = "Count")
  ,barmode = "group",
  
  updatemenus = list(
    
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.5, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[1:12]
    ),
    list(
      # buttons as the type of menu we want, as well as styling for buttons
      type = "buttons",
      direction = "down",
      x = 1.7, y = 1,
      showactive = FALSE,
      
      # object to be used for the buttons
      buttons = q2FinalGraphButtons[13:18]
    )
  )
)


# hides the tool bar that shows up on plot_ly interactive graphs
q2FinalGraph = q2FinalGraph %>% config(displayModeBar = FALSE)

q2FinalGraph
```





### Database

```{r}

library(DBI)
library(RSQLite)
drv = dbDriver("SQLite")

pokemonDatabase = dbConnect(drv, dbname = "./data/veekun-pokedex.sqlite")
dbExecute(pokemonDatabase, "PRAGMA foreign_keys = on")

```


```{r, eval = FALSE}
dbListTables(pokemonDatabase)

```

<a id="hypothesis-testing"></a>

### **Hypothesis Testing**

Pokémon types follow fairly standard themes, with Water types usually being fish, Flying types being birds, ect. Then the logical conclusion would be that the stats of these Pokémon reflect their physical forms, with Flying types being birds with high speed stats and larger Ground types being more durable but with lower speeds. But does this hypothesis hold up statistically?
<br>
<ul>
<li><div><p>**Null Hypothesis:**</p> <p>There is no statistically significant difference in the distributions of Flying type Pokémons' Speed stats and the Ground type Pokémons' Speed stats.</p></div></li>

<li><div><p>**Alternative Hypothesis:**</p><p>Flying type Pokémon have a statistically higher distribution of speed stats than Ground types.</p></div></li>
</ul>


<br>

**Step 1: Query -> data frame with every Flying type and their speed stat and every ground type and their speed stat**

**Step 1a: get a dataframe with every Pokemon and its speed stat**
```{r}
dbGetQuery(pokemonDatabase, 
           "PRAGMA FOREIGN_KEY_LIST(pokemon_stats)")

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, identifier, base_stat
           FROM pokemon_stats INNER JOIN stats
           ON pokemon_stats.stat_id = stats.id
           WHERE identifier = 'speed'
           LIMIT 20")

```


**Step 1b: dataframe with every pokemon's type**
```{r}
# dbGetQuery(pokemonDatabase, "
#            SELECT *
#            FROM pokemon_types")
# 
# dbGetQuery(pokemonDatabase, "
#            PRAGMA FOREIGN_KEY_LIST(pokemon_types)")
# 
# dbGetQuery(pokemonDatabase, "
#            SELECT *
#            FROM types")
dbGetQuery(pokemonDatabase, "
           PRAGMA FOREIGN_KEY_LIST(pokemon_types)
           ")  

dbGetQuery(pokemonDatabase, "
           SELECT pokemon_id, slot, identifier as TYPE
           FROM pokemon_types INNER JOIN types
           ON pokemon_types.type_id = types.id
          LIMIT 20")


```

**Step 1c: Dataframe with every Pokémon, its speed stat, and its type**
```{r, eval = FALSE, echo = FALSE}

dbGetQuery(pokemonDatabase, "SELECT * FROM pokemon_stats")

dbGetQuery(pokemonDatabase, "SELECT * FROM stats")

dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon_stats)")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(pokemon)")
dbGetQuery(pokemonDatabase, "PRAGMA FOREIGN_KEY_LIST(stats)")


```

```{r}
# initial query with all required columns
dbGetQuery(pokemonDatabase, "
           SELECT pokemon.id, pokemon.identifier, pokemon_types.pokemon_id, pokemon_types.type_id, types.identifier, pokemon_stats.stat_id, pokemon_stats.base_stat, stats.identifier
           
           FROM pokemon_types

           INNER JOIN pokemon ON pokemon_types.pokemon_id = pokemon.id
           INNER JOIN types ON pokemon_types.type_id = types.id
           INNER JOIN pokemon_stats ON pokemon_stats.pokemon_id = pokemon.id
           INNER JOIN stats ON pokemon_stats.stat_id = stats.id
           LIMIT 20
           ")

```


**Step 1d: Save the dataframes**
```{r}

# cleaned up query with all required info for flying and ground type pokemon
flynGrndSPD_df = dbGetQuery(pokemonDatabase, "
           SELECT pokemon.identifier AS name, types.identifier AS type, stats.identifier AS stat_name, pokemon_stats.base_stat
           
           FROM pokemon_types

           INNER JOIN pokemon ON pokemon_types.pokemon_id = pokemon.id
           INNER JOIN types ON pokemon_types.type_id = types.id
           INNER JOIN pokemon_stats ON pokemon_stats.pokemon_id = pokemon.id
           INNER JOIN stats ON pokemon_stats.stat_id = stats.id
            
           WHERE stat_name LIKE 'speed' AND (type LIKE 'flying' OR type LIKE 'ground')
           
           ")

# data frame with flying types and their speed stats
flyingSpeeds = flynGrndSPD_df[which(flynGrndSPD_df$type == "flying"), ]

# data frame with ground types and their speed stats
groundSpeeds = flynGrndSPD_df[which(flynGrndSPD_df$type == "ground"), ]

# first 10 entries
flynGrndSPD_df[1:10,]
```


**Step 2: Determine normality**
<br>
Visualize distribution using Bar Plots and QQ Plots 
```{r}
hist(flyingSpeeds$base_stat, main = "Flying Type Speed Distribution", xlab = "Speed stat", breaks=20)
```

```{r}
hist(groundSpeeds$base_stat, main = "Ground Type Speed Distribution", xlab = "Speed stat", breaks = 20)
```

```{r}
qqnorm(flyingSpeeds$base_stat, main  = "Normal Q-Q Plot for Flying Types")
qqline(flyingSpeeds$base_stat, col = "red")

```

```{r}
qqnorm(groundSpeeds$base_stat, main  = "Normal Q-Q Plot for Ground Types")
qqline(groundSpeeds$base_stat, col = "red")
```

**Step 2b: Conduct Shapiro-Wilk Test**

<br>
While the data doesn't appear to be normal, it helps to do a formal test for an accurate interpretation. Conducting a Shapiro-Wilk Test on a random sample from both groups will be more definitive.


<br>
First take random samples from both groups with sample size n = 30.
```{r}
set.seed(5423)
flySpeedSample = sample(flyingSpeeds$base_stat, 30, replace = TRUE)
grndSpeedSample = sample(groundSpeeds$base_stat, 30, replace = TRUE)
```

Then use R's built in function for the Shapiro-Wilk test for normality.
```{r}
shapiro.test(flySpeedSample)
shapiro.test(grndSpeedSample)
```

**Interpreting the results:**

<br>
The W statistic is the result of the Shapiro Wilk Test, which measures how  close the sample matches a normal distribution, with 1 being normal and 0 being not normal. 

<br>
The P value determines if the null hypothesis (that the random sample comes from a normally distributed data set) can be rejected. If the p value result is <= 0.05, null hypothesis is rejected. However if the p value result is > 0.05 you are unable to reject the null hypothesis.

<br>
However, depending on the random sample, the samples either fail or pass the test. To see if the samples are passing the Shapiro-Wilk test a reliable number of times, build a function that does the Shapiro-Wilk test on 100 random samples from the data. Using a p value of 0.05 (ie if less that 95% of the samples pass the test) to determine if the null hypothesis that the samples come from a normally distributed population can be rejected.

```{r}
shapiroSample = function(dataset, reps, sampleSize){
  

  pass_vec = c()
  for(i in 1:reps){
    sample = sample(dataset, sampleSize, replace = TRUE)
    outcome = shapiro.test(sample)
    
    # checks to see if this sample rejects the null hypothesis or not
    # if it does adds a 0 for fail, and if not adds a 1 for pass
    
    if(outcome[2] > 0.05){pass_vec[i] = 1}
    else if(outcome[2] <= 0.05){pass_vec[i] = 0}
    
  }
  
  return(pass_vec)
}

```


Execute the function
```{r}
set.seed(2354)
mean(shapiroSample(flyingSpeeds$base_stat, 100, 30))

mean(shapiroSample(groundSpeeds$base_stat, 100, 30))
```
Since the percentage of samples out of 100 that pass the Shapiro Wilk test are less than 95% for both flying and ground types, the null hypothesis can be rejected.
<br>

**Step 3: Conduct Mann-Whitney U Test**
<br>
The Mann-Whitney U test is the appropriate statistical test to utilize when comparing distributions of non-normal data.
<br>

**Step 3a: Rank the speed stats of both samples**
<br>
Showing each sample and the top 20 ranks
```{r, echo = FALSE}
# samples of flying that we used for the shapiro wilk test
flySpeedSample
grndSpeedSample

# creates a data frame with columns type and speed_stat and populates them with the samples
fly_grnd_speed_ranked = data.frame(rep(c("flying", "ground"), each = length(flySpeedSample)), c(flySpeedSample, grndSpeedSample) )
names(fly_grnd_speed_ranked) = c("type", "speed_stat")

# orders the speed_stat column(ascending)
fly_grnd_speed_ranked = fly_grnd_speed_ranked[order(fly_grnd_speed_ranked$speed_stat), ]

# adds a column with the ranks 
fly_grnd_speed_ranked$rank = 1:dim(fly_grnd_speed_ranked)[1]
fly_grnd_speed_ranked[c(1:20), ]
```

**Step 3b: Separate the flying and ground types to get their rank sums**
```{r}
ground_speed_ranks = fly_grnd_speed_ranked[which(fly_grnd_speed_ranked$type == "ground"), ]

flying_speed_ranks = fly_grnd_speed_ranked[which(fly_grnd_speed_ranked$type == "flying"), ]
```

**Step 3c: Calculate the rank sums for both types**
```{r}
ground_speed_ranksum = sum(ground_speed_ranks$rank)
flying_speed_ranksum = sum(flying_speed_ranks$rank)
ground_speed_ranksum
flying_speed_ranksum
```

**Step 3d: Determine the U statistic and the P value**

The U statistic is the measure of how often values from one of the samples rank below the other. U1 is the number of times values from the first sample rank above values from the second one. U2 is vice versa. The U statistic is min(U1, U2). 

<br>
However as sample sizes increase past n = 20, even though the U statistic still shows how much overlap there is between the two sample ranks, the distribution of U, ie all of the possible values that U could be, becomes approximately normal.

<br>
This requires calculating the z-score instead, the measure of how many standard deviations your calculated U statistic is from the mean U statistic for distributions under the null hypothesis.

<br>
Then use pnorm(z-score) to get the p value, the measure of how likely, given the null hypothesis is true, of getting that specific z-score. 
```{r}
n1 = dim(ground_speed_ranks)[1]
n2 = dim(flying_speed_ranks)[1]

U1 = ( (n1 * n2) + ((n1 * (n1+1)) / 2) ) - ground_speed_ranksum
U2 = ( (n1 * n2) + ((n1 * (n1+1)) / 2) ) - flying_speed_ranksum

U1
U2

Ustatistic = min(U1, U2)
muU = (n1 * n2) / 2
sigmaU = sqrt(((n1 * n2) * (n1 + n2 + 1)) / 12)
  
zscore = (Ustatistic - muU) / sigmaU

zscore

# gives the p value
pnorm(zscore)
```

Since the p value is extremely small (p < 0.01), much smaller than the chosen significance value of 0.05, the null hypothesis that there is no significant difference between the speed stats of flying and ground type Pokémon is rejected.

<br>
Additionally a negative z-score like the one above indicates that the first group(ground types) had lower ranks.
<br>

#### **Conclusion:**

After taking a a random sample of Flying and Ground type Pokémon, preliminary analysis using the Shapiro-Wilk test indicated that the samples for both types did not come from a normal distribution. As a result, the non parametric Mann-Whitney U test was applied, resulting in a z-score and p value that revealed a statistically significant difference in speed stats of Flying-type and Ground-type Pokémon. These results align with the original hypothesis and thematic characteristics of both types of Pokémon, indicating that there is an important relationship between a Pokémon's type and it's stats.


```{r}

dbDisconnect(pokemonDatabase)

```













































